The Evolution of Phenotypically Invariant Genetic Networks

### 2020-06-18 kalman_walk_paper.tex
added refs

%%%%%%%%%%%%%%%%%%%%%%
%\begin{multicols}{2}
\section*{Introduction}
  \jss{I'm not sure we still need this first paragraph -- I don't feel like it says that much?}
It is an overarching goal of many biological subdisciplines 
to attain a general understanding of the function and evolution of the 
complex molecular machinery that translates an organism's genome 
into the characteristics on which natural selection acts,
the phenotype.
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.
There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
and empirically inspired
\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.

 % In biological systems, the relationship between structure and function is often not invertible -- that is information about function is not usually sufficient to determine structure.
 % Non-invertiblility thus entails a diversity of functionally equivalent structures.
  Biological systems, at nearly all levels of organization, display considerable degeneracy, or (perhaps more descriptively) a capacity for isofunctional \emph{structural plasticity} -- 
  that is systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
  This is a consequence of the often non-invertibility (i.e. degeneracy) of the mapping between a system's structure and function.
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick, woesse}
all the way up to the convergent evolution of adaptive traits. Between these levels, structural plasticity can be found in the \emph{neutral networks}
  of nucleic acid sequences that map to an RNA secondary structure \citep{gruner, andmore}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide},
as well as in the thermodynamic stability of proteins \citep{hart2014thermodynamic}. Further examples include the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription} 
and neural circuits \citep{trojanowski}, and in developmental systems \citep{true2001developmental}. 
\revpoint{1}{1} \revpoint{1}{3}
  \jss{Other ideas instead of "degeneracy" or "isofunctional structural plasticity": non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}

@@

Structural plasticity has been argued to be necessary for natural selection \citep{edleman2001degeneracy}, 
as it may function as a mechanism for biological robustness and evolvability \citep{robustness}, or manifest as 
\emph{canalization} \citep{canalization}. Degeneracy of the genetic code, and of genetic networks, 
for instance, can make sequences more fault-tolerant to mutations and open up pathways to adaptation (cite). 
Non-invertibility can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura},
and perhaps even contribute to the formation of new species \citep{gavrilets}.

@@

We have focused on examples of single traits (where the phenotype is one-dimensional),
but phenotypes under selection are often high-dimensional,
and different traits often share a genetic basis.
Even in networks where the phenotype and kryptotype are of the same dimension,
system theory shows us that there will always be available degrees of freedom
as specific system realizations are only unique up to a change of coordinates.
Some phenotypes, however, require kryptotypic dimensions to be larger than that of the phenotype.
For instance, there exists minimal realizations (\emph{e.g.} the oscillator in Example 1) where the dimension of the kryptotpye is larger
than that of the phenotype, implying that for these phenotypic dynamics to be realized, the kryptotype dimension \emph{has} to be larger than the dimension of the phenotype.
Of course we see many more degrees of freedom in nonminimal networks, but it is not yet clear if evolution commonly arrives at such structures.
There are some hints that this may be occuring: compared to the molecular circuitry of the cyanobacteria circadian rhythm,
the human clock appears to be highly and unneccessarily complex \citep{sancar2008intelligent}. 
The true ``dimensionality of trait space'' has been debated at some length in the literature
\citep{trait_space}.
Whatever it is, however, it seems likely to be highly degenerate,
%Whatever it is, however, it seems unlikely 
%that it would approach the dimensionality of kryptotype space,
%which is something related to the number of genes.
even under universal pleiotropy \citep{universal_pleiotropy,omnigenic},
XXX

>>>>

%%%%%%%%%%%%%%%%%%%%%%
%\begin{multicols}{2}
\section*{Introduction}
It is an overarching goal of many biological subdisciplines 
to attain a general understanding of the function and evolution of the 
complex molecular machinery that translates an organism's genome 
into the characteristics on which natural selection acts,
the phenotype.
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.
There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
and empirically inspired
  \citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. \jss{Something, something Wagner model here?}

 % In biological systems, the relationship between structure and function is often not invertible -- that is information about function is not usually sufficient to determine structure.
 % Non-invertiblility thus entails a diversity of functionally equivalent structures.
  Biological systems, at nearly all levels of organization, display considerable degeneracy, or (perhaps more descriptively) a capacity for isofunctional \emph{structural plasticity} -- 
  that is systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
  This is a consequence of the often non-invertibility (i.e. degeneracy) of the mapping between a system's structure and function.
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick1963recent}
all the way up to the convergent evolution of adaptive traits. Between these levels, structural plasticity can be found in the \emph{neutral networks}
  of nucleic acid sequences that map to an RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide1997neutral},
as well as in the thermodynamic stability of proteins \citep{hart2014thermodynamic}. Further examples include the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription} 
and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{true2001developmental}. 
\revpoint{1}{1} \revpoint{1}{3}
  \jss{Other ideas instead of "degeneracy" or "isofunctional structural plasticity": non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}

@@

Structural plasticity has been argued to be necessary for natural selection \citep{edelman2001degeneracy}, 
as it may function as a mechanism for biological robustness and evolvability, or manifest as 
\emph{canalization} \citep{whitacre2010degeneracy}. Degeneracy of the genetic code, and of genetic networks, 
for instance, can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy} and open up pathways to adaptation \citep{wagner2008robustness}.
Codon degeneracy can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura1968genetic},
and perhaps even contribute to the formation of new species \citep{gavrilets2014models}.

@@

We have focused on examples of single traits (where the phenotype is one-dimensional),
but phenotypes under selection are often high-dimensional,
and different traits often share a genetic basis.
Even in networks where the phenotype and kryptotype are of the same dimension,
system theory shows us that there will always be available degrees of freedom
as specific system realizations are only unique up to a change of coordinates.
Some phenotypes, however, require kryptotypic dimensions to be larger than that of the phenotype.
For instance, there exists minimal realizations (\emph{e.g.} the oscillator in Example 1) where the dimension of the kryptotpye is larger
than that of the phenotype, implying that for these phenotypic dynamics to be realized, the kryptotype dimension \emph{has} to be larger than the dimension of the phenotype.
Of course we see many more degrees of freedom in nonminimal networks, but it is not yet clear if evolution commonly arrives at such structures.
There are some hints that this may be occuring: compared to the molecular circuitry of the cyanobacteria circadian rhythm,
the human clock appears to be highly and unneccessarily complex \citep{sancar2008intelligent}. 
The true ``dimensionality of trait space'' has been debated at some length in the literature
\citep{trait_space}.
Whatever it is, however, it seems likely to be highly degenerate,
%Whatever it is, however, it seems unlikely 
%that it would approach the dimensionality of kryptotype space,
%which is something related to the number of genes.
even under universal pleiotropy \citep{universal_pleiotropy, boyle2017expanded},
XXX
